Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Squeezing.

\begin{abstract}
There is growing interest in accelerating irregular data-parallel
algorithms on GPUs.  These algorithms are typically \emph{blocking},
so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
do not mandate fair scheduling, and GPU schedulers are inherently
unfair in practice.  This issue is currently circumvented by
exploiting scheduling quirks particular to today's GPUs in a manner
that does not allow the GPU to be shared with other workloads (such as
graphics rendering tasks).  We propose \emph{cooperative kernels}, an
extension to the traditional GPU programming model geared towards
writing blocking algorithms.  Workgroups of a cooperative kernel are
fairly scheduled, supporting blocking algorithms, and co-scheduling of
other tasks is supported via a small set of language extensions that
create a contract between the programmer and the scheduler.  We
describe the semantics of the cooperative kernels programming model
and discuss our portable prototype implementation.  We evaluate the
technique by porting a set of irregular work stealing and graph
algorithms to our programming model and assessing the performance and
responsiveness of the approach across several GPUs.  Because our
implementation exploits no vendor-specific hardware, driver or
compiler support, our encouraging experimental results provide a
lower-bound on the efficiency with which cooperative kernels can be
implemented in practice.

@@

\paragraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  The last decade has seen growing interest in accelerating irregular
data-parallel algorithms on GPUs, particularly algorithms that
traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.

@@

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling, which in the context of GPUs means fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and current GPUs do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
The simplest way for a GPU driver to handle 
the case where more workgroups are launched than there are compute
units is via an \emph{occupancy-bound}
execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16}.  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.   It is
evidently a simple model for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, which are
bound to separate compute units that operate in parallel.  Current GPU
implementations of irregular data parallel algorithms assume the
occupancy-bound execution model, which they exploit by launching 
no more workgroups than there are available compute
units~\cite{owens-persistent}. This works because \emph{today}'s GPUs seem to provide the
occupancy-bound execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable.
% from the point-of-view of
%GPU vendors and end users.
First, the execution model does not enable
multitasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.
%Precisely because
%of this problem, operating systems employ a GPU \emph{watchdog} to
%heuristically detect and kill long-running computations on the GPU
%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that (1) GPU vendors will
not commit to the occupancy-bound execution model that they currently
implement, due to the need for multitasking and energy throttling,
yet (2) GPU vendors will not guarantee fair scheduling, due to the
runtime overhead associated with supporting full preemption of
workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
retain the essence of the occupancy-bound model, which is
simple to implement efficiently, supporting preemption
only in key special cases.

\paragraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms, such as most irregular algorithms,
demand a degree of fair scheduling of workgroups, but for good reasons
GPU vendors will not commit to the guarantees of even the
occupancy-bound execution model.

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A GPU kernel
that requires workgroups to be fairly scheduled is identified as a
\emph{cooperative} kernel, and is written using two additional
language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates that the workgroup
would ideally continue with its execution but that, if required, the
scheduler may destroy the workgroup for good: the cooperative kernel
is prepared to deal with either scenario.

At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commence execution directly after the $\offerfork$ program point.

@@

\begin{enumerate}

@@

\item Cooperative kernels can be implemented over the occupancy-bound
  execution model that current GPUs provide---we present a
  proof-of-concept implementation that requires
  no special-purpose hardware or driver support.

@@

\end{enumerate}

@@

\begin{itemize}

@@

\end{itemize}

We begin by providing an overview of the OpenCL programming model and
two motivating examples of irregular algorithms (\mysec\ref{sec:background}).  At the end of the paper we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).

@@

GPUs are known to have relaxed memory models~\cite{ }.  To abstractly
account for this, we assume that the shared state component $\sigma$
is not simplyl a mapping from locations to values, but instead
captures all the intricacies of GPU memory that can lead to this
relaxed behaviour.  We also assume a function $\sync$ which, given a
kernel state $\kappa$, returns a set of kernel states.  The idea is
that each $\kappa' \in \sync(\kappa)$ is a possible kernel state that
can be reached from $\kappa$ by stalling until all stores to memory
and loads from memory to thread-local storage have completed.  All we
require is that $\sync$ does not modify the number of active
workgroups nor the component of a thread state that determines which
statements remain to be executed.

>>>>

\begin{abstract}
There is growing interest in accelerating irregular data-parallel
algorithms on GPUs.  These algorithms are typically \emph{blocking},
so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
do not mandate fair scheduling, and GPU schedulers are inherently
unfair in practice.  Current approaches avoid this issue by exploiting
scheduling quirks particular to today's GPUs in a manner that does not
allow the GPU to be shared with other workloads (such as graphics
rendering tasks).  We propose \emph{cooperative kernels}, an extension
to the traditional GPU programming model geared towards writing
blocking algorithms.  Workgroups of a cooperative kernel \emph{are}
fairly scheduled, and multitasking is supported via a small set of
language extensions that the kernel and scheduler use to cooperate.
We describe the semantics of the cooperative kernels programming model
and our prototype implementation.  We evaluate the approach by porting
a set of irregular work stealing and graph algorithms to our
programming model and examining their performance under multitasking
across three GPUs.  Our implementation exploits no
vendor-specific hardware, driver or compiler support, so our encouraging
results provide a lower-bound on the efficiency with
which cooperative kernels can be implemented in practice.

@@

\paragraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  There is growing interest in accelerating such algorithms on GPUs, particularly algorithms that
traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.

@@

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. The current GPU
programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and current GPUs do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice, depending on the kernel, multiple workgroups might map to the same compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle 
the case where more workgroups are launched than there are compute
units is via an \emph{occupancy-bound}
execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16}.  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.   It is
evidently a simple model for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, which are
bound to separate compute units that operate in parallel.  Current GPU
implementations of blocking algorithms assume the
occupancy-bound execution model, which they exploit by launching 
no more workgroups than there are available compute
units~\cite{owens-persistent}. This works because \emph{today}'s GPUs seem to provide the
occupancy-bound execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable.
% from the point-of-view of
%GPU vendors and end users.
First, the execution model does not enable
multitasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  The GPU cannot be used meanwhile for other
tasks, such as graphics rendering.
%Precisely because
%of this problem, operating systems employ a GPU \emph{watchdog} to
%heuristically detect and kill long-running computations on the GPU
%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
Second, \emph{energy throttling} is
an important concern for battery-powered devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down some compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that (1) for the above reasons, GPU vendors will
not commit to the occupancy-bound execution model that they currently
implement,
yet (2) GPU vendors will not guarantee fair scheduling, due to the
runtime overhead associated with supporting full preemption of
workgroups.  Vendors instead wish to
retain the essence of the simple occupancy-bound model, supporting preemption
only in key special cases.

\paragraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms
demand fair scheduling of workgroups, but for good reasons
GPU vendors will not commit even to the guarantees of the
occupancy-bound execution model.

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A kernel
that requires fair scheduling is identified as \emph{cooperative}, and written using two additional
language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates that the workgroup
would ideally continue executing, but that if required the
scheduler may destroy the workgroup for good.  The cooperative kernel
must be prepared to deal with either scenario.
%
At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commence execution directly after the $\offerfork$ program point.

@@

\vspace{-1mm}
\begin{enumerate}

@@

\item Cooperative kernels can be implemented over the occupancy-bound
  execution model that current GPUs provide: our proof-of-concept implementation requires
  no additional hardware or driver support.  Thus the model should be eay for vendors to implement directly.

@@

\end{enumerate}
\vspace{-1mm}

@@

\vspace{-1mm}
\begin{itemize}

@@

\end{itemize}
\vspace{-1mm}

We begin by providing background on OpenCL via two motivating examples (\mysec\ref{sec:background}).  At the end we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).

@@

GPUs are known to have relaxed memory models~\cite{ABDGKPSW-2015}.  To abstractly
account for this, we assume that the shared state component $\sigma$
is not simplyl a mapping from locations to values, but instead
captures all the intricacies of GPU memory that can lead to this
relaxed behaviour.  We also assume a function $\sync$ which, given a
kernel state $\kappa$, returns a set of kernel states.  The idea is
that each $\kappa' \in \sync(\kappa)$ is a possible kernel state that
can be reached from $\kappa$ by stalling until all stores to memory
and loads from memory to thread-local storage have completed.  All we
require is that $\sync$ does not modify the number of active
workgroups nor the component of a thread state that determines which
statements remain to be executed.
